Relation of severe COVID-19 in Scotland to transmission-related factors in individuals with and without risk conditions eligible for shielding support: REACT-SCOT case-control study

### 2021-02-10 shielding.Rmd
updated shielding markdown script before circulation

---
title: "Risk of COVID-19 in clinically extremely vulnerable individuals advised to shield during the epidemic: the REACT-SCOT case-control study" 
author:
  - name: Paul M McKeigue
    affiliation: "HPS,Usher"
  - name: David Caldwell 
    affiliation: HPS
  - name: Ciara Gribben
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Matthew Armstrong
    affiliation: HPS
  - name: Joke Delvaux
    affiliation: HPS
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: David McAllister 
    affiliation: Glasgow  
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Jim McMenamin  
    affiliation: HPS
  - name: Alison Smith-Palmer 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Helen M Colhoun
    affiliation: "HPS,IGMM,NHSFife"
  - name: Public Health Scotland COVID  -19 Health Protection Study Group
    affiliation: HPS

address: 
- code: IGMM
  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care  
- code: HPS
  address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
- code: GCU
  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
- code: ISD
  address: Information Services Division, NHS Scotland, UK

header-includes:
  \usepackage{newunicodechar}
  \DeclareUnicodeCharacter{039B}{\ensuremath{\mit\Lambda}}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
  \usepackage{lscape}
  \usepackage{pdfpages}
  \usepackage[nolists, nomarkers]{endfloat}

always_allow_html: true
urlcolor: blue
output:
      bookdown::pdf_book:  
        base_format: rticles::elsevier_article
        keep_tex: true
        fig_caption: yes

geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"

bibliography: ./covid.bib
csl: ./biomed-central.csl
linenumbers: false
linkcolor: cyan

@@

Prepared for the Shielding Policy Evaluation team and the Clinical Leads for Shielding Group (CLAGS)

Correspondence: Professor Helen Colhoun 07791652665/ helen.colhoun@igmm.ed.ac.uk

```{r functions, echo=FALSE}
library(knitr)
library(kableExtra)
library(bookdown)

options(knitr.kable.NA = '')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
source("helperfunctions.R")

@@

@@

# Background  
The SARS-CoV-2 pandemic reached Scotland in early 2020 with the first positive laboratory test being recorded on March 1st. Extreme social distancing measures were soon imposed, as in many other countries, with subsequent relaxation now that the effective reproduction number of cases is below 1. The overall policy is one of suppression and if monitoring indicates a second wave of COVID-19, more stringent measures are likely to be be re-imposed.

Early in the epidemic it was realised that certain people would likely be extremely vulnerable to severe COVID-19 disease including those receiving certain immunosuppressant drugs or with other conditions.  Therefore Public Health Scotland  instituted a process of identifying such persons through their primary care physician and through searching of administrative databases  (see appendix for full protocol). Those considered eligible then received a letter detailing the need for self isolation and the support they would receive to achieve this. These letters were sent out over several weeks between xx-and xx.

An initial survey of ~12000 shielding eligible persons indicates that in general some degree of compliance was high but often partial and it varied across different socio-demographic groups.

Here we take the original listing of ~ 180,228 persons deemed eligible for shielding at the outset of the epidemic. We examine the cumulative incidence of any evidence of COVID-19 and any serious COVID-19. using a case control design of all serious COVID-19 cases and age and sex matched controls in Scotland we examine the odds of disease associated with being shielding eligible. We explore whether there is any evidence that shielding reduced risks in these persons.

Using data from the period in which shielding was then stopped ( 1 August 2020) and the next instruction issued to take extra  precautions we examine the number of cases that have occurred.

# Methods
The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020-1].

## Ascertainment of eligibility for shielding

To identify  all those nationally deemed eligible for shielding we took the list generated by PHS at the outset of the epidemic (see https://hpspubsrepo.blob.core.windows.net/hps-website/nss/3008/documents/1_covid-19-search-criteria-highest-risk-patients.pdf  for details) .

In brief the following groups were identified

* Solid organ transplant recipients

* People with specific cancers

* People with severe respiratory conditions: cystic fibrosis, severe asthma, severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)

* People with rare diseases

* People on immunosuppression therapies

* People who are pregnant with significant heart disease, congenital or acquired.

* People on renal dialysis and who have had a splenectomy were also added ** to which group?

* A 7th group for primary and secondary care clinicians to identify patients who should be advised to shield.

 As noted in the search criteria document cited above patients were identified via searches of hospital discharge and prescribing records, as well as other sources including regional cancer chemotherapy databases, NHS Blood and Transplant and individual specialist clinician lists.

The list contained the CHI number which enabled linkage to other datasets. it also contained# the reason for eligibility and the date on which the shielding letter was sent to the person concerned.

## COVID-19 case ascertainment and definition
For the total Scottish population "Evidence of  any detected COVID-19" was obtained as having had a positive  nucleic acid test  for SARS-CoV-2 in  the national virology laboratory database,  or having had a hospital discharge code for COVID-19 in those with no positive test, or a death with COVID-19 mention anywhere on the death certificate even where testing had not been  positive.  The databases used were  the Electronic Communication of Surveillance in Scotland (ECOSS) database  that captures all virology testing in all NHS laboratories nationally,  the National Records of Scotland (NRS) death registrations data, the RAPID
database which is a daily update of hospitalisation nationwide and Scottish Morbidity Records-01 which captures hospital discharges  and discharge diagnostic codes nationally.

Databases were queried from March 1st 2020  (date of the first diagnosed case of COVID-19 in Scotland) up to the `r format(max(cc.all$SPECIMENDATE), "%d %b %Y")` for test results and `r format(max(cc.all$DATE_OF_REGISTRATION, na.rm=TRUE), "%d %b %Y")` for death registrations.

All these databases use the  Community Health Index (CHI) identifier. The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland. The CHI database is the master database for the CHI number and  contains several fields including age, sex, residential postcode, and residential care home status.

Among all those with any evidence of detected COVID-19 we then defined severe disease as that requiring critical care or  fatal.   All admissions to intensive (ICU), high dependency (HDU) or combined ICU/HDU care (hereafter critical care unit-CCU) were obtained by linkage using the CHI number  to the Scottish Intensive Care Society and Audit Group (SICSAG) database. Fatal COVID-19 was defined  based on their being a mention of  COVID-19 anywhere  on the death certificate regardless of prior testing or not, OR death within 28 days of testing positive  (or of the admission date for those SMR01 ascertained cases) regardless of whether COVID-19 was mentioned on the death certificate or not. Note that these definitions of COVID-19 deaths are the official ones used by the  National Records of Scotland and Public Health Scotland, respectively.

## Ascertainment of controls
For every case i.e. person with "any evidence of COVID-19" "as defined above, we selected 10 controls matched for sex, one year age band and attending the same GP practice  and alive on the same date of the index case , test, two weeks prior to index case admission or two weeks prior to death for cases ascertained through testing, SMR01 admission data, or NRS death data respectively.  The sampling frame for controls was the Community Health Care Index.

## Demographic data
For all those on the shielding list   and for all COVID-19 cases and controls i)socioeconomic status was evaluated using the postcode-based indicator the Scottish Index of Multiple Deprivation (SIMD) using the residential postcode in the CHI database [@fraser_scottish_2020]. Postcodes were allocated to which quintile of the socioeconomic distribution of Scottish residential datazones they belonged.  (ii) Ethnic group
was captured by linkage to prior SMR01 and SMR00 and SMR04  records   in which the
patients self assigns ethnicity according to the groups that are used in the
national census. iii) Residential care home status for those with diabetes was available via the
care home flag for postal address in the CHI database.

## Other morbidity and drug prescribing
For all those on the shielding list   and for all COVID-19 cases and controls, so as to ascertain other comorbidities the  general hospital admissions data in the  Scottish Morbidity Record 01 (SMR01),  and  all discharge codes over the past five years and Anatomical Therapeutic Chemical codes [@who_anatomical_2016] from Prescribing Information Systems (PIS) data for the past three years were extracted and used to define co-morbid conditions and prior drug exposures .

We first extracted history of a specific list of conditions and drug classes that have been included as moderate risk conditions for COVID-19 by public health agencies hereafter termed "listed conditions" [@nhs_whos_2020]. These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record, or a dispensed prescription in which that diagnostic code is the only indication (see Supplementary Materials for details).

## Statistical methods
\begin{itemize}

\item First we tabulated the shielding list by age-group  sex and reason  for shielding.

\item We then tabulated the \% ( and CI for the \%) of those on the shielding list with "Any evidence of COVID-19" as described above by age and sex   and reason for shielding as well as overall.

\item We also tabulated the \% of those on the shielding list with "Severe i.e. CCU treated or fatal COVID-19 " as described above, by age and sex   and reason for shielding as well as overall.

\item To compare risks in shielded persons before and after shielding was introduced For each shielded person we  marked up their longitudinal record  from Feb 7th with date of shielding letter, date of stop shielding letter (Aug 1st)  and any date of death with the record being censored at data of death. We then computed the number of cases of any evidence of COVID-19  per total  person days in pre shielded time, then shielded time , then post shielded time.  We repeated this for severe cases.

\item Among the shielding to understand whether there are other determinants of becoming a case, we compared the  \% with each of the  other listed conditions,  the \% with more than 1 listed condition, the number of prescribed ATC drug classes, the \% in  residential care home,  the  SIMD  distribution and ethnicity  distribution in those who became cases and those who did not.

\item To provide a comparator for the background population,  we used  the 2019 age band and sex - mid year estimates for the general population and the shielding status in cases to derive  the \% ( and 95\%CI) with any evidence of COVID-19 and severe COVID-19 in the same age and sex groups in the general population  (DAVID this is available from STUART). We then used logistic regression of the one year age band counts of cases and non cases in the shielded and general population to compute and OR associated with shielding.

\item The odds of "exposure "to eligibility for shielding in cases  versus  controls was also  quantified as the conditional odds ratio  from a conditional logistic regression.  We did this because it was easy to extend this regression was then extended to include SIMD, ethnicity, residential care home status,  the "listed "COVID-19 risk conditions.

\item Since the calculation in 4. above can be confounded by overall background trends in the Re to further evaluate if there is any evidence that shielding reduced risk, The odds of "exposure "to eligibility for shielding in cases  versus  controls  BY WEEK of epidemic was also quantified as the conditional odds ratio  from a conditional logistic regression.

\end{itemize}

Analyses were conducted using the R statistical package .

# Results

Table \ref{tab:shieldgroups} shows the frequency of shielding categories. The 6 women pregnant with heart disese were allocated to the "clinician identified" category.

Table \ref{tab:letterdates} shows a breakdown by date of sending letter

Table \ref{tab:afterletter} shows cases among shielded individuals classified by whether estimated date of infection was before or after the shielding letter.

Table \ref{tab:regression.allcases} shows the estimated rate ratios for any diagnosis of COVID-19

Table \ref{tab:regression.severecases} shows the estimated rate ratios for severe COVID-19.

# Discussion

## Summary of findings

## Comparison with previous literature

## Limitations and advantages 
Some key advantages of our study are the population coverage, the range of
outcomes assessed, the extensive prior risk factor and clinical history
data.

## Policy Implications

# Declarations
## Information governance

>>>>

---
title: "Relation of severe COVID-19 in Scotland to transmission-related factors in individuals with and without risk conditions eligible for shielding support: REACT-SCOT case-control study"
author:
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    affiliation: HPS
  - name: David McAllister\textsuperscript{\getAff{Glasgow}}  
    affiliation: HPS 
  - name: David Caldwell 
    affiliation: HPS
  - name: Ciara Gribben
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Matthew Armstrong
    affiliation: HPS
  - name: Joke Delvaux
    affiliation: HPS
  - name: Sam Colville
    affiliation: HPS
  - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}} 
    affiliation: HPS
  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}}   
    affiliation: HPS
  - name: Nazir Lone
    affiliation: Usher
  - name: Jim McMenamin 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    email: helen.colhoun@igmm.ed.ac.uk (+44 7791 652665)
    affiliation: HPS
address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care  
- code: HPS
  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
- code: GCU
  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
- code: IGMM
  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology

header-includes:
  \usepackage{newunicodechar}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
  \usepackage{lscape}
  \usepackage{pdfpages}
output:
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    keep_tex: true
    fig_caption: yes
#    number_sections: false
    md_extensions: -autolink_bare_uris
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
csl: ./biomed-central.csl
always_allow_html: true
urlcolor: blue
linenumbers: false
linkcolor: cyan

@@

```{r functions, echo=FALSE}
library(knitr)
library(kableExtra)

options(knitr.kable.NA = '.')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
knitr::opts_chunk$set(fig.width=6, fig.pos = "H", out.extra = "")
source("helperfunctions.R")

@@

dateformat <- function(x) { # format date and strip leading zero
     gsub("^0", "", format(x, "%d %B %Y"))
}

format.estcipv <- function(x.ci, x.pval) {
    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
    x <- gsub(", ", " to ", x)
    x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
    x <- gsub("times", "\\\\times", x)
    return(x)
}

peakdate <- casedates.allgr[which.max(avdaily), date]
maxdeaths <- casedates.allgr[which.max(avdaily), avdaily]
deaths.quartermax.date <- casedates.allgr[date > peakdate & avdaily < 0.25 * maxdeaths,
                                 date][1]

peakdate.elig <- casedates.byelig[listedgr3=="Eligible for shielding"][which.max(avdaily), date]
maxdeaths.elig <- casedates.byelig[listedgr3=="Eligible for shielding"][which.max(avdaily), avdaily]
deaths.elig.quartermax.date <- casedates.byelig[listedgr3=="Eligible for shielding" &
                                         date > peakdate.elig &
                                         avdaily < 0.25 * maxdeaths.elig,
                                         date][1]

r <- as.numeric(gsub(" .*", "", table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 5]));

pc <- gsub(".*\\(", "", table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 2])
pc <- 0.01 * as.numeric(gsub("\\\\.*", "", pc))

@@

 \newpage

# Abstract

* _Objectives_ - To investigate the relation of transmission-related factors-- occupation, household composition and recent hospital exposure -- to severe COVID-19 in  clinically vulnerable individuals eligible for shielding and in the general population.

* _Design_ - Matched case-control study (REACT-SCOT).

* _Setting_ - population of Scotland from 1 March 2020 to `r dateformat(lastdate.specimen)`.

* _Participants_ -  All `r nrow(cc.all[CASE==1])` diagnosed cases of COVID-19 and `r nrow(cc.all[CASE==0])` controls matched for age, sex and primary care practice, linked with all `r nrow(shielded.full)` individuals identified as eligible for shielding by Public Health Scotland.

* _Main outcome measure_ - severe COVID-19, defined as cases that entered critical care or were  fatal.

* _Results_ - With those without risk conditions as reference category, the univariate rate ratio for severe COVID-19 was
`r format.estcipv(table.listedgr3.severecases.nocare["Moderate risk condition", 3], table.listedgr3.severecases.nocare["Moderate risk condition", 4])` 
in those with moderate risk conditions and
`r format.estcipv(table.listedgr3.severecases.nocare["Eligible for shielding", 3], table.listedgr3.severecases.nocare["Eligible for shielding", 4])` 
in those eligible for shielding.  The highest risk for severe disease was in solid organ transplant recipients: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])` .   In all risk groups severe COVID-19 was strongly associated with recent hospital exposure: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`  overall. The fraction of severe cases attributable to recent hospital exposure peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% at the beginning of May 2020, and again at `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  at the beginning of December 2020. Risk of severe COVID-19 increased with the number of adults in the household but decreased with the number of school-age children in the household.  The rate ratio associated with two or more adults (with single-adult households as reference category), was   `r format.estcipv(table.listedgr3.severecases.nocare["adultsgt1", 5], table.listedgr3.severecases.nocare["adultsgt1", 6])` and the rate ratio associated with one or more school-age children was `r format.estcipv(table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 5], table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 6])`.

* _Conclusions_ -  Mitigating the impact of the epidemic requires control of transmission in hospital settings.  Shielding of vulnerable individuals until they have been vaccinated requires protection from transmission from other adults in the same household.  For solid organ transplant recipients, who are at very high risk and in whom the efficacy of vaccines is uncertain, vaccination should be offered to household contacts.

# Background  
The SARS-CoV-2 pandemic reached Scotland in early 2020 with the first positive laboratory test recorded on 1 March 2020. Population-wide interventions included advice that symptomatic individuals should self-isolate, banning of mass gatherings, closure of schools and finally a population-wide lockdown on 23 March 2020.  Although no systematic studies of risk factors were available at that time, public health agencies issued lists of "moderate risk" conditions [@nhs_whos_2020] and "diseases and conditions considered to be very high risk" [@kanani_caring_2020].  Those with "very high risk" conditions were designated as clinically extremely vulnerable and eligible for shielding.

The shielding programme issued individual letters to eligible persons advising on the need for social distancing, recommending strict isolation and giving information on the provided support which included a national opt-in scheme of free food delivery, home delivery of medication and priority access to supermarket delivery slots. Whether the programme had an impact on the infection rates in those shielded is unclear.

The initial objective of this study was to investigate  whether the shielding letter and programme had any demonstrable impact on the risk of COVID-19 in those eligible for shielding  versus those with moderate risk or the rest of the population.  For this we hypothesized that  a fall in  the risk ratio  for those eligible for  shielding  after the receipt of shielding letters  would be evidence of impact. Our second objective was to understand the role of  transmission-related factors: household composition, recent hospital exposure including out-patient consultations, and exposure related to occupations including health care [@shah_risk_2020] and teaching [@mcallister_report_2020]  on COVID-19 among those eligible for shielding.    As the relevance of these transmission-related factors among the shielded  became clear, the objectives were broadened to investigation of the relation of severe COVID-19 to these transmission-related factors in  the remainder of the population also.

# Methods
In brief; we linked a dataset of all those eligible for shielding  nationally to an existing case-control dataset (REACT-SCOT) conducted by PHS that includes all COVID-19 cases in Scotland and matched population controls.[@mckeigue_rapid_2020-1]   This case control dataset is refreshed every few weeks and  through  linkage to health and other records already contained extensive covariate data on a wide range of risk factors and settings for COVID-19 and derived variables for   "moderate risk conditions". Thus all cases and controls were  classifiable into three categories: no risk condition, moderate risk condition only, eligible for shielding.  The rate ratio of COVID-19 associated with each risk category was estimated  as the conditional odds ratio  from the case-control study as described below.

## Ascertainment of eligibility for shielding 
We used the dataset of those eligible for shielding that was created by PHS for the purposes of issuing letters.   This dataset comprised a list of  patient identifiers and the shielding condition and the date the shielding letter was issued. The list had been generated  by querying health-care information systems including hospital discharge records, prescription encashments, regional cancer chemotherapy databases, blood and transplant registries,  for evidence of a list of agreed diseases and conditions  and supplemented by data submitted  to PHS following direct requests to clinicians in relevant specialties [@health_protection_scotland_search_2020].

The categories designated as eligible for shielding were as listed below [@health_protection_scotland_search_2020; @kanani_caring_2020]:

1. Solid organ transplant recipients
2. People with cancer of the blood or bone marrow at any stage of treatment, or people with cancer receiving treatments that affect the immune system 
3. People with severe respiratory conditions including cystic fibrosis, severe asthma and severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)
4. People with rare diseases that increase the risk of infections such as severe combined immunodeficiency, homozygous sickle cell disease. 
5. People on immunosuppression therapies sufficient to increase risk of infection
6. Women who were pregnant with heart disease
7. Other patients identified by clinicians as requiring shielding advice. 
8. Those on renal dialysis and those who had had a splenectomy (included from 11 May 2020)

The dataset of  those eligible for shielding process  has been regularly refreshed.  Here we used the dataset comprising  `r nrow(shielded.full)` individuals identified up to `r dateformat(lastdate.specimen)`. The first batch of  shielding letters was sent on `r dateformat(batches$Date.Sent)[1]`.   Further batches were issued on a weekly basis and the programme was paused on  1 August 2020. On 25 November 2020 a further letter was issued with "extra protection level advice for people at highest risk" based on the current protection level for the population level in that area.

## Ascertainment of cases and sampling of controls
Case ascertainment for the REACT-SCOT study has been described in detail elsewhere [@mckeigue_rapid_2020-1].  Case ascertainment was based on querying the following national-level databases: Electronic Communication of Surveillance in Scotland (ECOSS) that captures virology testing in all NHS laboratories, National Records of Scotland (NRS) death registrations,  RAPID which is a daily update of hospitalisations, and Scottish Morbidity Record 01 (SMR01) which records general hospital discharges including day cases and is ICD-10 coded. All these databases use the Community Health Index (CHI) number as identifier.  The CHI database includes age, sex, postcode and care home status, and can be queried to extract numbers of adults and children in the household.

Cases of COVID-19 were defined as those with a positive  nucleic acid test for SARS-CoV-2 in ECOSS, a hospital discharge diagnosis of COVID-19 in SMR01, or a death registration with mention of COVID-19 anywhere on the death certificate. The presentation date was assigned as the date of the first positive test for those ascertained through testing, as seven days before the admission date for those without a positive test result ascertained through hospital discharge records, and as fourteen days before the date of death for those without a positive test result ascertained through death certificates. Databases were queried from 1 March 2020  (date of the first diagnosed case of COVID-19 in Scotland) up to `r dateformat(maxdate.specimen)` for test results and `r dateformat(maxdate.death)` for deaths.

Severe COVID-19 was defined by entry to critical care within 21 days of presentation date, or fatal outcome. Entry to critical care units -- intensive care, high dependency or combined units -- was obtained by linkage to the Scottish Intensive Care Society and Audit Group (SICSAG) database. Fatal outcome was defined as death at any time with COVID-19 coded as underlying cause, death from any cause within 28 days of testing positive or death within 28 days of presentation date for cases ascertained through discharge records.  This definition of severe cases was chosen to ensure that case ascertainment would not be affected by triage of those assessed as unlikely to benefit from critical care, or by the scale-up of SARS-CoV-2 testing.

For each case of COVID-19, up to 10  community controls matched for sex, one-year age band and primary care practice were selected from the CHI database, and assigned the same presentation date as the case. With this incidence density sampling design, it is possible and correct for an individual to appear more than once as a control and subsequently as a case.

## Linkage of cases and controls to demographic and morbidity data
Linkage of cases and controls to demographic and morbidity data and the associations of these factors with severe COVID-19 have been described in detail elsewhere [@mckeigue_rapid_2020-1].  Cases and controls were linked to hospital discharge ICD-10 codes over the last five years in SMR01, to British National Formulary codes of dispensed prescriptions in the 240 days before presentation date in the Prescribing Information System (PIS) and to the national register of diabetes.  We used these linked datasets to assign a list of "moderate risk conditions" for COVID-19 designated by public health agencies [@nhs_whos_2020]: diabetes, heart disease, asthma or chronic airway disease, chronic kidney disease, disabling neurological conditions, and immune deficiency or suppression.  The codes used are as described previously  [@mckeigue_rapid_2020-1].  Three risk groups were defined: no risk condition, moderate risk condition but ineligible for shielding, and eligible for shielding.  Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation (SIMD) score which is based on linkage of postcodes to Census data [@fraser_scottish_2020].  Quintile 5 is the least deprived.

# Transmission-related risk factors 
Through linkage of the CHI database to the Unique Property Reference Number it was possible to calculate the numbers of adults and children in each household and to refine the care home residence  assignation contained in the CHI database.  Care home residence was assigned using the field in the CHI database, augmented by coding as care home residents the `r care.home.reclassified[1]` individuals aged over 70 in households with 10 or more adults to give a total of `r cc.all[care.home=="Care/nursing home", .N]` care home residents out of the `r cc.all[, .N]` cases and controls.  Linkage to occupational status for health-care workers and teachers was undertaken as described elsewhere [@shah_risk_2020; @mcallister_report_2020].

We used the Scottish Morbidity Records SMR01 and SMR00 (outpatient attendance) to examine possible associations with exposure to health-care facilities. We defined the variable "recent hospital exposure" as any hospital in-patient stay or face-to-face out-patient consultation from 5 to 14 days before presentation date. Restriction of hospital exposure to this time window was intended to exclude consultations caused by COVID-19 symptoms for which testing was delayed by a few days, but to include those exposed to health care facilities in the time interval during which the infection was likely to have been acquired.  Our intention was to capture all cases that *could have been* nosocomial rather than the more restrictive definition  used in Scotland and England for "probable healthcare -associated infection" as illness onset (or first positive specimen date at least 8 days after admission  which aims to  identify cases that are unlikely to have been infected outside hospital [@public_health_england_covid-19_2020-1].

## Statistical methods
The analyses presented here focus  mainly on severe COVID-19 as this case definition does not depend upon the extent of testing for SARS-CoV-2, which varied markedly over the study period and is also likely to differ between those eligible for shielding and those ineligible.  Care home residents were excluded from case-control analyses as shielding advice intended for individuals in private households would not have been relevant to care home residents.

Cases and controls were grouped into three categories: no risk condition, moderate risk condition but ineligible for shielding, and eligible for shielding. Rate ratios for severe COVID-19 associated with eligibility for shielding are calculated with "no risk condition" as reference category. All rate ratios were estimated by conditional logistic regression.  As the  community controls were drawn by incidence density sampling and matched for age, sex, and general practice, this controls for these variables and for calendar time.

To plot the time course of the rate ratios associated with eligibility for shielding and with hospital exposure, these rate ratios were estimated over `r winsize`-day sliding windows of calendar time.  The fraction of severe cases attributable to hospital exposure was calculated in each 21-day time window from Miettinen's formula as \ensuremath{p_c \left( r - 1 \right) / r}.  where \ensuremath{p_c} is the frequency of exposure in cases in that time window and \ensuremath{r} is the rate ratio [@hsieh_effect_1988].  Results were similar with use of Levin's formula \ensuremath{p \left( r - 1 \right) / \left[ 1 + p \left( r - 1 \right) \right]}  where \ensuremath{p} is the frequency of exposure in controls.

Sliding windows of size  `r winsize.casedates`, `r winsize.hosp`, and `r winsize.fatality` days were used to plot the number of severe cases, the frequency of recent hospital exposure, and the rate ratios associated with hospital exposure and household composition.  In the plots of exposure frequency and rate ratio the data points from 1 June 2020 to 30 September 2020 are omitted as the numbers of cases and controls are too low for estimates of frequencies and rate ratios to be accurate.

# Results
A methodological concern was that unrecorded filtering of fatalities while the list of those eligible for shielding was being prepared could have led us to misclassify early fatal cases among those eligible for shielding as ineligible, leading to underestimation of the rate of fatal disease in the first few weeks of the epidemic in those eligible for shielding. As described in the Supplementary Material, by examining the time course of the fatal case proportion we were able to establish that such misclassification was unlikely in cases presenting after the second week of March 2020.

## Shielding eligibility and risk of COVID-19
Table \ref{tab:shieldgroupsbyage} shows the frequency of risk categories in those eligible for shielding.  To comply with statistical disclosure control rules, the `r shielded.full[shield.group=="Pregnant with heart disease", .N]` women who were identified as in the category "pregnant with heart disease" have been omitted from this table and all subsequent analyses; there were no severe cases in this group.

The largest risk category was those with severe respiratory disease.   Solid organ transplant recipients and those on immunosuppressants were younger than those in the other categories. Supplementary Table \ref{tab:letterdates} shows the proportions in each of these risk categories by date of advice letter: most of the solid organ transplant recipients, those with severe respiratory disease, and cancer patients were included in the first batch of letters on April 3rd 2020, but most of those in other risk categories were included in later batches.

Table \ref{tab:shieldcasegroup} shows  among the shielding group the number in each eligibility category by case status classified as not diagnosed, not severe or severe COVID-19.

Overall among `r nrow(shielded.full)` persons eligible  for shielding `r table.shield.casegr["All shielding categories", 2]` were diagnosed with COVID-19 but did not enter critical care or die within 28 days, `r table.shield.casegr["All shielding categories", 3]` entered critical care for COVID-19 but survived, and `r table.shield.casegr["All shielding categories", 4]` had a fatal outcome. Those on immunosuppressants had the lowest risks.

Figure \ref{fig:plotcasesbydate} compares the time series of severe cases in the three risk groups as a ribbon plot.  The daily number of presentations of severe cases in the total population reached a peak on `r dateformat(peakdate)` and fell to one quarter of the peak by `r dateformat(deaths.quartermax.date)`.   After each peak, the number of daily cases fell more slowly in those eligible for shielding than in those without risk conditions.

## Rate ratio for severe COVID-19  by risk condition group   
Of those eligible for shielding who were included in the case-control dataset, `r cc.severe[CASE==1 & shield.any==TRUE, .N, by=care.home][care.home=="Care/nursing home", N]` of the `r cc.severe[CASE==1 & shield.any==TRUE, .N]` severe cases and `r cc.severe[CASE==0 & shield.any==TRUE, .N, by=care.home][care.home=="Care/nursing home", N]` of the `r cc.severe[CASE==0 & shield.any==TRUE, .N]` matched controls were resident in care homes.  In all subsequent case-control analyses care home residents and their matched controls have been excluded, as policies for shielding residents of care homes are different from those relevant to shielding individuals living independently.

Table \ref{tab:regressionseverecasesnocare} shows the rate ratios for severe COVID-19 associated with shielding eligibility and with presence of a moderate risk condition, with those without risk conditions as reference category.  The univariate rate ratio for severe disease was
`r format.estcipv(table.listedgr3.severecases.nocare["Moderate risk condition", 3], table.listedgr3.severecases.nocare["Moderate risk condition", 4])` 
in those with moderate risk conditions and
`r format.estcipv(table.listedgr3.severecases.nocare["Eligible for shielding", 3], table.listedgr3.severecases.nocare["Eligible for shielding", 4])` 
in those eligible for shielding.  Among those eligible for shielding, solid organ transplant recipients were the group at highest risk, with a univariate rate ratio of  `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])` for severe COVID-19.     On multivariate adjustment rate ratios for all these groups remained high.

### Time course of the rate ratio associated with eligibility for shielding 
Figure \ref{fig:plotrateratio} (upper panel) shows the time course of the univariate rate ratios for severe COVID-19 n the  shielding and moderate risk categories, with those without risk conditions as reference category.   These rate ratios are computed over `r winsize`-day windows of presentation dates.  The rate ratio associated with eligibility for shielding increased from `r round(coeffs.long[date.midpoint==as.Date("2020-04-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 April to `r round(coeffs.long[date.midpoint==as.Date("2020-05-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 May.  Therefore there was no evidence that  the  shielding letters reduced the rate ratio associated with shielding eligibility.  A similar trend of rate ratio over time was found for those with moderate risk conditions.
